Speeding up decimal multiplication

### 2020-11-15 paper/paper.tex
Update

\subsubsection{Montgomery reduction}

@@

It is possible to compute $\redc$ efficiently; the algorithm internally
(not counting the raw $xy$ multiplication)
performs:
\begin{itemize}
    \item one multiplications of $w$ bits by $w$ bits into $(2w)$ bits;
    \item one multiplications of $w$ bits by $w$, of which only the lowest $w$ bits are used.
\end{itemize}
This is more efficient than the Barrett reduction if one does not consider the cost of conversion
into and out of Montgomery representation. In the case of number-theoretic transform, we can
convert everything into Montgomery representations at the beginning (this is $\Theta(N)$ time),
then perform the transform (this is $\Theta(N \log N)$ modular multiplications), and then
convert everything out of Montgomery representation (this is, again, $\Theta(N)$ time).

@@

\subsection{Omitting the bit-reversal permutation}

@@

@@

@@

@@

Remember the algorithm for computing $\redc^\ast(t) = R^{-1} t$
in $\mathrm{F}_p,$
assumes $$0 \le t < p \mu,$$ where $\mu = 2^w$ if the machine word is $w$ bits long.
We can then compute $\redc(x,y),$ where $0 \le x,y < p$, as $\redc^\ast(xy).$

Remember also that we require $p < \mu/2.$

>>>>

\subsubsection{Montgomery reduction}
\label{section:montgomery_intro}

@@

It is possible to compute $\redc$ efficiently. We now give the definition of
the procedure $\mathrm{MontgomeryReduce}(a).$
If we represent field elements as values in $[0; p-1],$ then
$\redc(x,y)$ can be computed as $\mathrm{MontgomeryReduce}(xy).$

\begin{procedure}[H]
\DontPrintSemicolon
\caption{MontgomeryReduce($a$)}
\KwData{
    Integer $\mu.$\;
    Integer $p$ coprime with $\mu.$\;
    Integer $p' \in [0; \mu-1]$ such that $p \cdot p' \equiv -1 \, (\mymod \mu).$
}
\KwIn{
    Integer $a \in [0; \mu p - 1].$
}
\KwOut{
    Integer $b \in [0; p-1]$ such that $b \equiv a \mu^{-1} \, (\mymod p).$
}
\Begin{
    $m \longleftarrow ((a \mymod \mu) \cdot p') \mymod \mu$\;
    $r \longleftarrow (a + m \cdot p) \mydiv \mu$
        \tcc*[r]{this division is exact} % apparently no need to use \; after \tcc
    \eIf{$r \ge p$}{
        \Return{$r - p$}\;
    }{
        \Return{$r$}\;
    }
}
\end{procedure}

The value of $p'$ can be pre-calculated with extended Euclidean algorithm for GCD;
we have $\gcd(p, \mu) = 1,$ so $\exists A, B \in \mathbb{Z} \colon A p + B \mu = 1.$
This means $A p \equiv 1 \, (\mymod \mu)$ and $-Ap \equiv -1 \, (\mymod \mu),$ so
$$p' = (-Ap) \mymod \mu.$$

The procedure $\mathrm{MontgomeryReduce(a)}$ internally
(not counting the raw multiplication $xy = a$)
performs:
\begin{itemize}
    \item one multiplications of $w$ bits by $w$ bits into $(2w)$ bits;
    \item one multiplications of $w$ bits by $w$, of which only the lowest $w$ bits are used.
\end{itemize}
This is more efficient than the Barrett reduction if one does not consider the cost of conversion
into and out of Montgomery representation. In the case of number-theoretic transform, we can
convert everything into Montgomery representations at the beginning (this is $\Theta(N)$ time),
then perform the transform (this is $\Theta(N \log N)$ modular multiplications), and then
convert everything out of Montgomery representation (this is, again, $\Theta(N)$ time).

@@

\subsection{Omitting the bit-reversal permutation}
\label{section:omitting1}

@@

Define $N = \log_2 |A|.$

Note that, in both DIT and DIF, we perform the same computation over and over again when doing
$$\varphi \longleftarrow \varphi \cdot \omega_m,$$
where $\omega_m = \omega^{2^{N - s}}, s = \log_2 m.$

One way to improve that is to pre-calculate powers of $\omega$ into a table $t,$ so that
$t[i] = \omega^i, 0 \le i < |A|/2.$
Then, given $j$ and $m=2^s,$ we can express the factor $\varphi$ as $t[j \cdot 2^{N - s}].$
Note that the multiplication by a power of two can be replaced with a bit shift.

Another way is to pre-calculate a separate table for each ``granularity'' of root powers:
define $$T_s[i] = t[i \cdot 2^{N-s}] = (\omega^{2^{N-s}})^i, 0 \le i < 2^{s-1}$$
for each $s, 1 \le s \le N.$ Note $T_N = t$ and $\sum\limits_{s=1}^{N} |T_s| = \sum\limits_{s=1}^{N} 2^{s-1} = 2^N-1.$
Now, given $j$ and $m=2^s,$ we can express $\varphi$ as $T_s[j].$

Let us write $T_s$ for all $s$ into a single array $T_\ast$
of length $2^N-1 = |A|-1$ so that
$$T_s[i] = T_\ast[(2^0 + 2^1 + \cdots + 2^{s-2}) + i] = T_\ast[2^{s-1} - 1 + i].$$
Then, given $j$ and $m=2^s,$ the factor $\varphi$ can be expressed as
$T_\ast[\Delta_s + j],$ where $\Delta_s = 2^{s-1}-1 = m/2 - 1.$

At first sight, there is not much difference between indexing
$T_\ast[C_1 + j]$ and $t[j \cdot 2^{C_2}]$
in the innermost loop,
where $C_1$ and $C_2$ denote loop-invariant expressions that can be calculated once before the loop.
But on x86 and x86-64, there is difference:
these platforms can encode SIB (scale-index-base) directly in the instruction that dereferences the
pointer, but only for constant scales of 1, 2, 4 and 8 bytes.
This means that, if our field element is 8, 16, 32 or 64 bits long, we can dereference $p[j],$
where $p$ points to $T_\ast[C_1],$ in one instruction; but dereferencing
$t[j \cdot 2^{C_2}],$ where $C_2$ can be anything, requires two instructions: first for the bit
shift and the second for dereferencing.

@@

Note that, in both DIT and DIF, the factor $\varphi$ for the step $j=0$ is $\widehat{1},$
the multiplicative identity of the field.
In this case, both butterflies degenerate into
$\langle x, y \rangle \mapsto \langle x+y, x-y \rangle.$

We can peel the first iteration off the loop to avoid redundant multiplication by $\widehat{1}.$

@@

Remember in section~\ref{section:omitting1} we established that, with
$x \cdot y$ denoting scalar (element-wise) product between $x,y \in \mathcal{F}^N,$
for any permutation $\pi$ we have $$\pi^{-1}(\pi(x) \cdot \pi(y)) = x \cdot y.$$

We then used this property to get rid of bit-reversal permutations in DIT and DIF.
It turns out that, for both four-step and six-step algorithms,
the same property can be employed to get rid of the
final transposition step in DFT and the initial transposition step in IDFT simultaneously.

For the four-step algorithm, this means no bit-reversal permutation or matrix transposition is ever
needed: we can also simultaneously omit the bit-reversal permutation steps in DIF and DIT
if we use DIF for the direct transforms in definition of four-step DFT, and DIT for inverse
transforms in the definition of four-step IDFT.

For the six-step algorithm, unfortunately, things are not so simple.
We can simultaneously omit the final transposition step (step 6) in the direct transform
and the initial transposition step (step 1) in the inverse transform; and can use DIF without
bit-reversal permutations for step 5 of the direct transform and DIT without bit-reversal
permutations for step 2 of the inverse transform; but we still have two matrix transposition steps
per transform, and no matter whether we use DIT or DIF for step 2 of the direct transform
and step 5 of the inverse transform, we need to perform bit-reversal permutation.

@@

Remember the procedure $\mathrm{MontgomeryReduce}(a)$ that we defined in section~\ref{section:montgomery_intro}.
